---
title: "Reproducible_research_assignment1"
author: "Christopher Stewart"
date: "February 12, 2015"
output: html_document
---

## Introduction 
This project presents a cursory analysis of some "quantified self" data for the first assignment in the Reproducible Research course in the Coursera Data Science Specialization. 

The data come for a personal activity monitoring device that collects data at 5-minute intervals. Two months worth of data consisting of the number of steps taken in these 5-minute intervals are presented. There are 17,568 observations with 3 variables: steps, date and interval.

Initially, we download and inspect the data.

## Loading and Preparing the Data for Exploratory Analysis

```{r download data, echo=TRUE}
require(downloader); require(Hmisc); require(plyr); require(lattice)
data_url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download(data_url, dest = "data.zip", mode = "wb")
unzip("data.zip", exdir = "./")
data = read.csv("activity.csv", stringsAsFactors = FALSE)
data$date <- as.Date(as.character(data$date))
data$interval <-as.factor(data$interval)
summary(data)
```

## First question: What is mean total number of steps taken per day?

For this portion of the analysis, we loop over the steps variable with "tapply" to calculate the mean number of steps per day. We then portray this in a histogram. Finally, we use the "summary" function to retrieve the median and mean 


```{r exploratory data analysis: 1, echo=TRUE}
stepday <- tapply(data$steps, data$date, mean)
print(stepday)
hist(stepday, main = "Total Number of Steps / Day", xlab = "Mean number of steps")
summary(stepday)
```

## Second question: What is the average daily activity pattern?

The second portion performs a similar analysis, but over intervals instead of days. Here we use the "aggregate" function to average steps over intervals. We then plot this using a basic time series plot and query the created object created to find the interval with the highest mean number of steps. The maximum mean steps per interval occurs at interval 835, which is the 104th interval.

```{r exploratory data analysis: 2, echo=TRUE}
stepint <- aggregate( data$steps ~ data$interval, FUN = mean )
plot(stepint, xlab = "Intervals from 0 to 2355", ylab = "Steps", type = "l", main = "Mean Number of Steps by Interval")
maxstepsint <- tapply(data$steps, data$interval, na.rm = TRUE, mean)
which.max(maxstepsint)
```


## Third question: Imputing missing values

The third analysis asks us to first compute the number of NAs in the data set. We then replace these values with the mean steps value in a new data frame using the "impute" function from the Hmisc package. To analyze the combined data, we create a new data frame that includes the imputed missing values with the rest ofthe data. 

We then loop through the vector with the imputed missing values to calculate the mean number of steps per day. In order to see the effect of imputation, we report the mean and median values of the vector with NAs vs. the vector with imputed missing values. Notice that the mean is identical, which is to be expected given that the imputed values were equal to the vector's mean, but the medians are different.

We then plot this as a histogram alongside the histogram without the imputed values to more clearly see the effect of missing data imputation on the two distributions.


```{r exploratory data analysis: 3, echo=TRUE}
length(which(is.na(data[1]==T)))
df.imp <- with(data, impute(data$steps, mean))
df.new <-data.frame(data$steps,data$date,data$interval,df.imp)
stepday2 <- tapply(df.new$df.imp, df.new$data.date, mean)
mean(stepday, na.rm = TRUE); median(stepday, na.rm = TRUE)
mean(stepday2); median(stepday2)
#create plots
par(mfrow=c(1,2))
hist(stepday, main = "Total Number of Steps / Day", xlab = "Mean number of steps")
hist(stepday2, main = "Imputation of NAs", xlab = "Mean number of steps")
```


## Fourth question: Are there differences in activity patterns between weekdays and weekends?

The fourth and final analysis asks us to use the "weekdays" function to make a panel plot depicting the changes in the "steps" variable from weekdays to weekends. 

To do so, we first use the "weekdays" function to determine whether dates are weekdays or weekend days. We then group days using the "revalue" function, transform that vector into class "factor" and, finally, aggregate mean step values over intervals and weekdays vs. weekend days. 


``` {r, exploratory data analysis: 4, echo=TRUE}
wkdays <-weekdays(df.new$data.date)
df.new$wkdays <- revalue(wkdays, c(Monday = "Weekday", Tuesday = "Weekday", Wednesday = "Weekday", Thursday = "Weekday", Friday = "Weekday", Saturday = "Weekend", Sunday = "Weekend"))
df.new$wkdays <-as.factor(df.new$wkdays)
stepint <-aggregate(df.imp ~ data.interval + wkdays, data = df.new, mean)
xyplot(df.imp ~ data.interval | wkdays, stepint, type = "l", layout = c(1, 2), xlab = "Intervals from 0 to 2355", ylab = "Steps", main = "Mean Number of Steps per Interval")
```